# Comparing k-means scenarios

Now that I've completed versions of the k-means process with AND without PCA as a data reduction step, I'd like to compare the models.
In the reading I did over the course of Jan 10-13, 2023, I found some good examples of other papers where they do this:

| 

Lasantha, V., Oki, T., & Tokuda, D.
(2022).
Data-Driven versus Köppen--Geiger Systems of Climate Classification.
*Advances in Meteorology*, *2022*, e3581299.
<https://doi.org/10.1155/2022/3581299>

Fovell, R. G., & Fovell, M.-Y. C.
(1993).
Climate Zones of the Conterminous United States Defined Using Cluster Analysis.
*Journal of Climate*, *6*(11), 2103--2135.
<https://doi.org/10.1175/1520-0442(1993)006%3C2103:CZOTCU%3E2.0.CO;2>

Green, P. E., & Krieger, A. M.
(1995).
A comparison of alternative approaches to cluster-based market segmentation.
*Market Research Society. Journal.*, *37*(3), 1--19.
\|

```{r setup}
#| message: false
#| echo: false
#| warning: false

library(tidyverse)
library(gt)

d_orig <- read_csv("data/mukey_cluster_assignments_and_soilprops.csv") %>% 
  select(mukey, k_6, k_8, k_11) %>% 
  mutate(across(.cols = contains("k_"), ~str_replace(.x, "Cluster_", "c")))

d_pca <- read_csv("data/pca_mukey_cluster_assignments_and_soilprops.csv") %>% 
  select(
    mukey,
    pk_6 = k_6, 
    pk_8 = k_8,
    pk_11 = k_11
  ) %>% 
  mutate(across(contains("pk_"), ~str_replace(.x, "Cluster_", "p")))

d <- left_join(d_orig, d_pca, by = "mukey")

```

## Which clusters are we comparing?

**Original k-means:** silhouette suggests 6, 9-11.
C-H suggests 6, 11.
Pairwise comparisons suggest 6, 8, or 11

**PCA before k-means:** silhouette suggests 4, 6, 8, 11.
C-H 4-8 would be similar, 9-11 would be similar

## Compare model metrics

```{r}
#| message: false

orig_metrics <- read_csv("data/kmeans_cluster_metrics.csv") %>% 
  mutate(version = "original")

pca_metrics <- read_csv("data/pca_kmeans_cluster_metrics.csv") %>% 
  mutate(version = "pca")

all_metrics <- bind_rows(orig_metrics, pca_metrics)

```

### Sum of within-cluster SSE

It doesn't make sense to plot these two on the same graph, because the errors are in different units (z-score for the original, PCA score for the PCA).

```{r}
#| echo: false

all_metrics %>% 
  ggplot(aes(x = n_clust, y = tot_wss, group = version)) + 
  geom_point(aes(color = version)) + 
  geom_line(aes(color = version)) +
  theme_bw() + 
  ylab("Sum of Within-Cluster SSE") + 
  xlab("Number of Clusters (k)") +
  ggtitle("Comparing Sum of Within-Cluster SSE") +
  labs(subtitle = "Lower values indicate that observations (MUKEYs) are closer\nto their cluster centroid (more similar to other MUKEYs in their cluster)") +
  theme(plot.subtitle = element_text(size = 9)) +
  scale_x_continuous(breaks = c(2:20)) + facet_wrap(vars(version))

```

### Average Silhouette

From the `{tidyclust}` documentation:

> Another common measure of cluster structure is called the **silhouette**.
>
> The silhouette of a single observation is proportional to the average distance from that observation to within-cluster observations minus the average distance to outside-cluster observations; normalized by the greater of these two average.
>
> In principle, a large silhouette (close to 1) suggests that an observation is more similar to those within its cluster than those outside its cluster.

```{r}
#| echo: false

all_metrics %>% 
  ggplot(aes(x = n_clust, y = avg_sil, group = version)) + 
  geom_point(aes(color = version)) + 
  geom_line(aes(color = version)) +
  theme_bw() + 
  ylab("Average Silhouette") + 
  xlab("Number of Clusters (k)") +
  ggtitle("Comparing Average Silhouette") + 
  labs(subtitle = "Higher is better, possible values [-1,1]") +
  theme(plot.subtitle = element_text(size = 9)) +
  scale_x_continuous(breaks = c(2:20))

```

### Comparing Calinski-Harabasz index

Higher values are better.
This is also known as the "Variance Ratio Criterion,", which is how Calinski & Harabasz refer to it in their 1974 paper introducing it.
The paper title is "A dendrite method for cluster analysis"

Nice summary [from PyShark](https://pyshark.com/calinski-harabasz-index-for-k-means-clustering-evaluation-using-python/):\

"The Calinski-Harabasz index (also known as the Variance Ratio Criterion) is calculated as a ratio of the sum of inter-cluster dispersion and the sum of intra-cluster dispersion for all clusters (where the dispersion is the sum of squared distances).

A high CH means better clustering since observations in each cluster are closer together (more dense), while clusters themselves are further away from each other (well separated)."

```{r}
#| echo: false

all_metrics %>% 
  ggplot(aes(x = n_clust, y = ch_index, group = version)) + 
  geom_point(aes(color = version)) + 
  geom_line(aes(color = version)) +
  theme_bw() + 
  ylab("Calinski Harabasz Index") + 
  xlab("Number of Clusters (k)") +
  ggtitle("Comparing Calinski Harabasz Index") + 
  labs(subtitle = "Higher is better") +
  theme(plot.subtitle = element_text(size = 9)) +
  scale_x_continuous(breaks = c(2:20))

```

## Jaccard similarity coefficient

So what am I comparing?
Visually, I'd like to make an alluvial plot or Sankey-type plot to see how the two compare (and I did this, see PDFS in `_refs/`. But I'd also like to compare membership between the different clusterings. In Lasantha et al., (2022), they use the Jaccard similarity coefficient.

> "The Jaccard similarity coefficient is the ratio between the intersection and union of two sets; it has values ranging from zero for non-intersection to one for exact similarity. This is index is widely used in the evaluation of similarity in clustering in addition to applications such as image recognition and text analysis" *Lasantha et al., 2022*

This appears to be the citation for the original coefficient:

P. Jaccard, "The distribution of the flora in the alpine zone zone" New Phytologist, vol.
11, no. 2, pp. 37--50, 1912.

From Tan et al. (2018): "... the simple matching coefficient, which is known as the Rand statistic in this context, and the Jaccard coefficient are two of the most frequently used cluster validity measures."

Additional relevant Jaccard citation:

Hennig, C.
(2007).
Cluster-wise assessment of cluster stability.
*Computational Statistics & Data Analysis*, *52*(1), 258--271.
<https://doi.org/10.1016/j.csda.2006.11.025>

Tan, P.-N., Steinbach, M., Karpatne, A., & Kumar, V. (2018).
*Introduction to data mining* (2nd ed.).
Pearson.

Equation:

Jaccard coefficient = f11 / (f01 + f10 + f11)

|                 | Same Cluster | Different Cluster |
|-----------------|--------------|-------------------|
| Same Class      | f11          | f10               |
| Different Class | f01          | f00               |

So in words, when we are comparing to sets of classes (like two different clusterings produced by k-means), the Jaccard coefficient is telling us the ratio of : objects that are in both sets (the intersection) divided by the union (total number of objects in both sets, subtracting the number they share)

Apparently the `{vegan}` package can do this with the `vegdist()` function, `method = 'jaccard'` .

The thing I didn't understand at first was that the Jaccard index is for use on a specific **pair of clusters**.
So I need to do this pairwise for all the clusters in the original and PCA versions

### Example from online tutorial

Jaccard index example from [UC Riverside GEN 242 Course](https://girke.bioinformatics.ucr.edu/GEN242/tutorials/rclustering/rclustering/#jaccard-index-for-cluster-sets)

```{r}

source("R/cindex_tgirke.R") 

library(cluster)

y <- matrix(rnorm(5000), 1000, 5, dimnames=list(paste("g", 1:1000, sep=""), paste("t", 1:5, sep="")))

clarax <- clara(y, 49)

# length = 1000, 49 classes (numeric 1:49)
clV1 <- clarax$clustering

clarax <- clara(y, 50)

clV2 <- clarax$clustering 

ci <- cindex(clV1=clV1, clV2=clV2, self=FALSE, minSZ=1, method="jaccard")



ci[2:3] # Returns Jaccard index and variables used to compute it

  n_intersect <- length(intersect(clV1, clV2))
  
  jac_index <- n_intersect/(length(clV1) + length(clV2) - n_intersect)

```

### My Jaccard function

This was a [helpful blog post](https://www.r-bloggers.com/2021/11/how-to-calculate-jaccard-similarity-in-r-2/) (Jaccard Index is quite simple to calculate).

Recall that what I want here is the MUKEYs.
That allows us to calculate an intersection (so how many MUKEYs appear in both clusters?).

```{r}

jaccard <- function(orig_clust, pca_clust, dat, orig_col, pca_col){
  
  orig_sym <- rlang::sym(orig_col)
  pca_sym <- rlang::sym(pca_col)
  
  orig_members <- d %>% 
    filter(!!orig_sym == orig_clust) %>% 
    pull(mukey)
  
  pca_members <- d %>% 
    filter(!!pca_sym == pca_clust) %>% 
    pull(mukey)
  
  n_intersect <- length(intersect(orig_members, pca_members))
  
  jac_index <- n_intersect/(length(orig_members) + length(pca_members) - n_intersect)
  
  return(jac_index)
}


```

Try it for k=6, calculating the Jaccard similarity coefficient for Cluster 1 from the original model set and Cluster 1 from the PCA model set.

To see if the result I'm getting makes sense, I'm looking at the alluvial diagram I made comparing original clusters and the PCA clusters.
What I see is that Cluster 1 from the original version is split between Cluster 1 and Cluster 2 in the PCA version, with roughly 2/3 of the MUKEYs shared between Cluster 1 (original) and Cluster 1 (PCA).
So it makes sense that the value below is 0.68.

```{r}

jaccard(orig_clust = "c1", 
        pca_clust = "p1",
        dat = d,
        orig_col = "k_6",
        pca_col = "pk_6"
        )


```

## Calculate Jaccard similarity coefficients

I think it makes the most sense to do this with three dataframes, one for each value of K we are interested in evaluating.
So that would be: 6, 8, 11.

### Set up data structure

```{r}

# data structure for k=6
clust_orig_k6 <- unique(d$k_6)
clust_pca_k6 <- unique(d$pk_6)

k6_pairs <- tidyr::crossing(k6_orig = clust_orig_k6,
            k6_pca = clust_pca_k6)

# data structure for k=8
clust_orig_k8 <- unique(d$k_8)
clust_pca_k8 <- unique(d$pk_8)

k8_pairs <- tidyr::crossing(k8_orig = clust_orig_k8,
            k8_pca = clust_pca_k8)

# data structure for k=11
clust_orig_k11 <- unique(d$k_11)
clust_pca_k11 <- unique(d$pk_11)

k11_pairs <- tidyr::crossing(k11_orig = clust_orig_k11,
            k11_pca = clust_pca_k11)
```

### Map over Jaccard function

```{r}

jac_k6 <- k6_pairs %>% 
  mutate(jaccard_coef = map2_dbl(.x = k6_orig,
                                 .y = k6_pca,
                                 .f = jaccard,
                                 dat = d,
                                 orig_col = "k_6",
                                 pca_col = "pk_6"),
         k = "k_6") %>% 
  rename(orig = k6_orig,
         pca = k6_pca)

jac_k8 <- k8_pairs %>% 
  mutate(jaccard_coef = map2_dbl(.x = k8_orig,
                                 .y = k8_pca,
                                 .f = jaccard,
                                 dat = d,
                                 orig_col = "k_8",
                                 pca_col = "pk_8"),
         k = "k_8") %>% 
  rename(orig = k8_orig,
         pca = k8_pca)

  
  
jac_k11 <- k11_pairs %>% 
  mutate(jaccard_coef = map2_dbl(.x = k11_orig,
                                 .y = k11_pca,
                                 .f = jaccard,
                                 dat = d,
                                 orig_col = "k_11",
                                 pca_col = "pk_11"), 
         k = "k_11") %>% 
  rename(orig = k11_orig,
         pca = k11_pca)


jac_all <- bind_rows(jac_k6, jac_k8, jac_k11)

write_csv(jac_all, "data/jaccard_coefficients_k6_k8_k11.csv")

```

### Visualize Results

```{r}
#| echo: false

jac_k6 %>% 
  select(-k) %>% 
  pivot_wider(names_from = "pca",
              values_from = "jaccard_coef") %>% 
  mutate(orig_label = "Full-6 Clusters") %>% 
  gt(rowname_col = "orig",
     groupname_col = "orig_label") %>% 
  data_color(
    columns = contains("p"),
    colors = scales::col_numeric(
      palette = colorRamp(c("#FFFFFF", "#31A354"),
                          interpolate="spline"),
      domain = c(0:1)
    ),
    apply_to = "fill"
  ) %>% 
  fmt_number(
    columns = contains("p"),
    decimals = 3
  ) %>% 
  tab_spanner(
    columns = contains("p"),
    label = "PCA-6 Clusters"
  ) %>% 
  tab_options(row_group.as_column = TRUE)
  
```

```{r}
#| echo: false

jac_k8 %>% 
  select(-k) %>% 
  pivot_wider(names_from = "pca",
              values_from = "jaccard_coef") %>% 
  mutate(orig_label = "Full-8 Clusters") %>% 
  gt(rowname_col = "orig",
     groupname_col = "orig_label") %>% 
  data_color(
    columns = contains("p"),
    colors = scales::col_numeric(
      palette = colorRamp(c("#FFFFFF", "#31A354"),
                          interpolate="spline"),
      domain = c(0:1)
    ),
    apply_to = "fill"
  ) %>% 
  fmt_number(
    columns = contains("p"),
    decimals = 3
  ) %>% 
  tab_spanner(
    columns = contains("p"),
    label = "PCA-8 Clusters"
  ) %>% 
  tab_options(row_group.as_column = TRUE)

```

```{r}
#| echo: false

jac_k11 %>% 
  select(-k) %>% 
  pivot_wider(names_from = "pca",
              values_from = "jaccard_coef") %>% 
  mutate(orig_label = "Full-11 Clusters") %>% 
  gt(rowname_col = "orig",
     groupname_col = "orig_label") %>% 
  data_color(
    columns = contains("p"),
    colors = scales::col_numeric(
      palette = colorRamp(c("#FFFFFF", "#31A354"),
                          interpolate="spline"),
      domain = c(0:1)
    ),
    apply_to = "fill"
  ) %>% 
  fmt_number(
    columns = contains("p"),
    decimals = 3
  ) %>% 
  tab_spanner(
    columns = contains("p"),
    label = "PCA-11 Clusters"
  ) %>% 
  tab_options(row_group.as_column = TRUE)

```
